Detection of eco-evolutionary dynamics in metacommunities using Joint Species

### 2024-11-11 manuscript/manuscript_v11.Rmd

@@

@@

>>>>

### Applying HMSC on empirical data

To test the applicability of HMSC on ecological and evolutionary effects in empirical data, we applied HMSC on the experimental data of Jewel & Bell (2023) [@Jewell2023]. The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* here designated as Lm, *Spirodela polyrhiza* designated as Sp and *Wolffia columbiana* designated as Wc, and the liverwort *Ricciocarpos natans* designated as Rn) under different environmental conditions for twelve weeks. The environmental conditions were defined by three differed amount of shadings of the water pool, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 μg L−1, dissolved phosphorous (DP) = 10 μg L−1)), medium (DN = 800 μg L−1, DP = 40 μg L−1), and high concentration (DN = 3200 μg L−1, DP = 160 μg L−1) of nutrients. They combined each shading with each nutrient level, resulting in nine different environmental conditions. Throughout the experiment they measured biweekly, from week one to week eleven, the frond size of ten individuals from each species, as body size trait, in each environmental condition alongside their corresponding species density. We calculated the average trait change as $\Delta x$ = $\^{x}_{ijt} - \^{x}_{ijt-1}$, where $\^{x}_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$. As no $\Delta \hat{x}$ value could be calculated for week one of the experiments, it was omitted from the analysis.  
We fitted two HMSC models to the data: An 'evolution model' in which $\Delta \hat{x}$ of each species and environmental conditions were applied as explanatory variable, and a 'no-evolution model', in which only the environmental conditions were included. Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model. Additionally, we applied replicates (*n* = 2) and time as random effects. For each model we ran four Markov chain Monte-Carlos' for 1,875,000 iterations from which the first 625,000 were cut off as the transient phase and with a thinning factor of 5,000 250 samples were taken from each chain.  
To test the impact of evolution by the model, we extracted the posterior distributions of the estimated beta effects. These were multiplied with the $\Delta \hat{x}$ of the corresponding species and summed up together with the intercept.

@@

### Empirical data

The empirical data [@Jewell2023] showed the community dynamics of four plant species under different environmental conditions, additionally the front size of each species was measured to determine its trait change over time. We calculated the mean variance partitioning (VP) over all species, to evaluate the impact of evolutionary (change of the front size trait) and ecological (light and nutrient treatments) effects on changes on species density. Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $\Delta \hat{x}$ (\@ref(fig:fig5)a). This is due to the reduction of the VP of the ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig5)). While $\Delta \hat{x}$ of the species Lm, Sp and Wc showed a similarly low impact, on average 3.6%, $\Delta \hat{x}$ of species Rn showed a high impact on variance of species density, 7.8%, which can be explained by its high $\Delta \hat{x}$ changes over time (see Appendix S05). The high density intervals (HDI) of 95% estimated beta effects showed a negative correlation of Light shading with species density, while nutrient concentration correlated positively with density (\@ref(fig:fig5)c). Changes in $\Delta \hat{x}$ impacted densities differently. At one hand changes of species Rn frond size correlated negatively with is density, on the other hand changes in the frond size of species Wc showed a positive correlation with the other species Sp and Lm (\@ref(fig:fig5)c). The impact of evolution (of all species summed up), impacted species density over time, yet did not show a clear direction over the course of the experiment (\@ref(fig:fig5)d).

```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments. The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents. The (b) high density interval of 95% showed a clear pattern for the different environmental treatments, yet differing results for trait changes and their impact on species density. This can be seen in (d) the shifting predictions of density over time due to the total trait changes of all species."), echo=FALSE}
knitr::include_graphics("../output/fig5.pdf")
```

@@

### Empirical data

By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to infer the relative impact of the trait changes on the community dynamics. This application shows is advantages over previous methods,which estimate the impact of trait changes (thus evolution), as it is not depended on the data structure and simplicity [@Hairston2005] [@Ellner2011], can be implemented on multiple species[@Lasky2020] and does not require additional common garden experiments [@Garzon2019]. Therefore this allows for a straight forward implementation of HMSC on a variety of data structures as lone as a clear response variable (e.g. species density) and ecological (e.g. environmental conditions) as well as evolutionary (e.g. heritable trait change) are measured. An additional advantage of HMSC, and JSDMs in general, is the ability to use the empirical data to predict a species response to changes in the environmental variables, even beyond the measured values [@Tikhonov2020] [@Ovaskainen2015]. With the implementation this approach can be combined with the fixed effects to answer further questions, such as: How does trait change impact the species density under changing environmental conditions? And how does the impact of trait changes depend on specific environmental conditions?

At the same time a correct interpretation of such predictions as well es the estimated posterior beta distribution require a thorough understand of the experimental system. For example we often observed a negative correlation between trait change and species density in our simulations and also in species Rn of the empirical data, yet they derived from different mechanisms. In our simulations trait changes were greater when a species trait was further away from the optimal trait value, this greater distance resulted also in lower fitness and thus lower population size, while a species with higher fitness and thus higher population size had smaller trait changes. Therefore, HMSC correlates a high trait changes with a low population size and thus estimates a negative beta effect. Such effects can be expected in empirical data of i.e. evolutionary rescue experiments, where a species has to adapt to a deleterious environment before going extinct [@Gomulkiewicz1995][@Bell2009]. In the empirical data this negative correlation was the result of the species Rn investing into a larger frond size (Fig. SX), this likely lead to a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.
